---
title: "IPUMS 2022 ACS Analysis"
author: 
- Dhruv Gupta
- Justin Klip
- Kevin Shen
abstract: The IPUMS data service was used to extract data regarding college graduates by state, and doctoral degrees holders by state. We use the Laplace estimation technique to obtain an estimate doctoral degrees by state, given that there were 391,171 respondents in California. We compare our estimates for states we assume education levels are relatively high (Massachussets, New York), to states that are perceived to be uneducated (Arkansas, Alabama), and compare our estimates to actual data from IPUMS. We find that highly educated states, like Massuchusetts are overestimated, likely because of the high number of PhDs in their population relative to California. Further, we find that for uneducated states like Arkansas, we underestimate the number of respondents.
bibliography: references.bib
date: today
date-format: long
format: pdf
---

```{r}
#| include: false
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(readr)
library(kableExtra)

# Load data into RStudio
data <- read_csv("/Users/dhgu2021/Downloads/usa_00001.csv")

```

## Instructions to Obtain the Data

In order to acquire the data used for our IPUMS research study, we first navigated to (https://www.ipums.org/). Once on the webpage, click on the "Visit Site" button under IPUMS USA. Next, click on the white "Get Data" button under CREATE YOUR CUSTOM DATA SET.

Once on the webpage used for selecting samples and variables, click on the blue "Select Samples" button on the left. On the USA Samples tab, uncheck the "Default sample from each year" checkbox, and then reselect the checkbox corresponding to 2022. Once selected, click the "Submit Samples" button; this will take you to the variable selection webpage.

On this page, navigate to the "Household" dropdown menu under Select Harmonized Variables, and select "Geographic." In the list of variables under Geographic Variables — Household, click the plus sign next to the "STATEICP" variable. Then, hover over the "Person" dropdown menu and select "Education." In the list of variables under Education Variables — Person, click the plus sign next to "EDUC."

In the Data Cart floating window, select "View Cart." On the Data Cart webpage, select the blue "Create Data Extract" button. On the Extract Request webpage, to the right of Data Format, click "Change" and switch the format to .csv. Finally, press the "Submit Extract" button at the bottom of the page.

On the Download or Revise Extracts webpage, find the extract you requested, and in the Download Data column, click "Download .csv." This will give you the data necessary to run the code for our analysis.

Following the steps above, we obtained geographical data regarding college graduates by state, and doctoral degrees in that state from the IPUMS data service [@ipums2022]. The analaysis for this data is done using [@citeR], and the packages used are Tidyverse, [@tidyverse] and KableExtra, [@kableExtra]. 

## Overview of Ratio Estimator Approach

We try to use the ratio estimators' approach to find total respondents for each state, given the fact that there are 391,171 respondents in California at all levels of education. We look at the number of total doctoral degree holders in California and compare it to the total number of respondents in California. We find there are about 61.74 times as many people compared to doctoral degrees in California. So, we simply multiply the number of doctoral degrees holders in each state by this ratio (61.74) to get our estimate of total respondents per state. The assumption is that other states would have similar ratios of PhD’s to total respondents. This approach is necessary if we only have doctoral data for all the states, but not total respondent data.


## Comparison of Estimated Totals and Actual Totals

@tbl-laplace compares our estimates of doctoral degree holders with actual totals for relevant states. The full table, @tbl-full, be found in the appendix. We display Massachussets (StateICP = 3), New York (StateICP = 13), Alabama (StateICP = 41), Arkansas (StateICP = 42), Wyoming (StateICP = 68), and California (StateICP = 71).

```{r}
#| include: false
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(readr)
library(kableExtra)

# Filter data for California (STATEICP == 71)
california_data <- data %>% filter(STATEICP == 71)

# Total number of respondents in California
total_respondents_california <- 391171  # Given value

# Number of respondents with doctoral degrees in California
doctoral_in_california <- california_data %>% filter(EDUCD == 116) %>% nrow()

# Calculate the ratio of doctoral degree holders to total respondents in California
ratio_california <- doctoral_in_california / total_respondents_california

# Group data by state and count respondents with doctoral degrees in each state
doctoral_by_state <- data %>%
  filter(EDUCD == 116) %>%
  group_by(STATEICP) %>%
  summarise(doctoral_count = n())

# Estimate the total number of respondents in each state using the ratio from California
estimated_total_by_state <- doctoral_by_state %>%
  mutate(Estimated_Total_Respondents = doctoral_count / ratio_california)

# Save the estimated data to a CSV file
write_csv(estimated_total_by_state, "estimated_total_by_state.csv")

```


```{r}
#| include: false
#| warning: false
#| message: false
actual_by_state <- data %>%
  group_by(STATEICP) %>%
  summarise(Actual_Total_Respondents = n())

# Merge the estimated data with the actual data
comparison <- estimated_total_by_state %>%
  left_join(actual_by_state, by = "STATEICP")

# Save the comparison data to a CSV file
write_csv(comparison, "comparison.csv")

```

```{r}
#| echo: false
#| label: "tbl-laplace"
#| warning: false
#| message: false

# Read the comparison data from the CSV file
comparison_data <- read_csv("comparison.csv")
# Round numerical columns to 2 decimal places
comparison_data <- comparison_data %>%
  mutate(
    doctoral_count = round(doctoral_count, 2),
    Estimated_Total_Respondents = round(Estimated_Total_Respondents, 2),
    Actual_Total_Respondents = round(Actual_Total_Respondents, 2)
  )

# Select specific states for the sample table
comparison_table_sample <- comparison_data %>%
  filter(STATEICP %in% c(68, 71, 13, 3, 42, 41)) %>%
  knitr::kable(
    caption = "Sample Comparison of Estimated and Actual Total Respondents for Selected States",
    col.names = c("State ICP", "Doctoral Count", "Estimated Total Respondents", "Actual Total Respondents")  # Custom column names
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Display the sample comparison table
comparison_table_sample

```

### Explanation as to Differing Results:

Some states may get differing values due to differing state characteristics. We based our ratio on California, which has a large number of educational institutes that award PhD’s. California’s large academic and advanced labor market may also draw in a lot of PhD’s in comparison to a state like Wyoming, which has less incentive for PhD’s to live there. This differing state level trends may explain why we get close estimates for some states, but not for others. In Wyoming (Code 68)  we predict there to be a much smaller number of people (4445.125) than there actually are (5962) and this is because of this discrepancy in assuming there are much more PhD’s in Wyoming then there are, so we multiply by too high of a number. Other states also underestimate suggesting the same issue as Wyoming. Massachusetts (Code 03), has an overestimate of the number of people, this is likely because they have a higher ratio of PhD graduates in their total population. This makes sense given that they have some of the best research institutions in the world that attract PhD’s, so in Massachusetts our ratio is an underestimate.

\newpage

# Appendix
```{r}
#| echo: false
#| label: "tbl-full"
#| warning: false
#| message: false

# Display the complete comparison table using kable
complete_comparison_table <- knitr::kable(
  comparison_data,
  caption = "Complete Comparison of Estimated and Actual Total Respondents by State",
  col.names = c("State ICP", "Doctoral Count", "Estimated Total Respondents", "Actual Total Respondents")  # Custom column names
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Display the complete comparison table
complete_comparison_table

```

# Bibliography